---
title: "lab03"
author: "Ryan McNally"
date: "9/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages('readxl')
dat <- readxl::read_xls('../ds202_lab3/GSS.xls')
library(ggplot2)
library(tidyverse)
library(dplyr)
getwd()
dat <- dat%>% 
  rename(year = 'Gss year for this respondent', happiness = 'General happiness', sex = 'Respondents sex', education = 'Highest year of school completed', marital = 'Marital status', id = 'Respondent id number', ballot = 'Ballot used for interview', partyid = 'Political party affiliation', income = 'Total family income')
```

1) Find the average years of education for the respondents with each marital status. Arrange your output in a meaningful order and print. Describe any patterns you find.

```{r}
dat$education<- suppressWarnings(as.numeric(dat$education))
dat%>%
  select(marital, education)%>%
  filter(education<21)%>%
  group_by(marital)%>%
  summarise(educ_mean = mean(education), .groups='drop')
```
**As shown in the chart above, the group with the highest average education is 'No answer'. The group with the lowest average education level is 'Widowed'. All of the groups are within the range of 11-15.**



2. Create a single data frame containing records for the better educated respondents with each marital status. A “better educated” respondent is someone who has strictly more years of education than the average among those with the same marital status. Print the structure of the data frame.
```{r}
df <- dat%>%
  group_by(marital)%>%
  filter(education<21)%>%
  mutate(maritalMedian = median(education))%>%
  filter(education > maritalMedian)%>%
  select(-maritalMedian)%>%
  ungroup()
str(df)
```
**To make this df, I first grouped the data by marital status like before. Next, I filtered by the education level. Next, I added a column that represented the median years of education for each education level.**


3. How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.
```{r}
dat%>% 
  select(happiness, marital)%>% 
  ggplot(aes(x=marital, fill=happiness)) + geom_bar(position='fill') + ggtitle("Happiness and Marital Status") + labs(x = "Marital Status", y = "Happiness Level")
dat%>% 
  select(happiness, marital)%>% 
  group_by(marital)%>%
  count(happiness)
```
**As shown by the graph above, it is clear to see that the group with the highest overall happiness is 'Married' people. The group with the lowest happiness is 'No answer'. The next lowest is the 'Separated'**



4. Does party affiliation affect the relationship you found in the last question?
```{r}
happy<-filter(dat, happiness %in% c('Very happy','Pretty happy'))
happy$marital<-factor(happy$marital, levels=(c("Married","Never married","Divorced","Widowed","Separated","No answer")))
hapState<-filter(dat, happiness %in% c('Not too happy','Very happy','Pretty happy'))
hapState$marital<-factor(hapState$marital, levels=(c("Married","Never married","Divorced","Widowed","Separated","No answer")))
ggplot(happy, aes(x=marital, fill=partyid))+geom_bar(position='fill')+ggtitle("Percent of Happy Respondents and Their Marital Status")+xlab("Marital Status")+ylab("Percent of Happy Respondents")
ggplot(hapState, aes(x=marital, fill=happiness))+geom_bar(position='dodge')+facet_wrap(~partyid)+ggtitle("Happiness vs. Marital Status")+xlab('Marital Status')+ylab('Number of Respondents')+coord_flip()
dat%>%
  select(happiness, marital, partyid)%>%
  group_by(marital, partyid)%>%
  count(happiness)
```
**From the graphs shown above, I can not see any obvious trends between political party affiliation and happiness.**

5. Is marital status related to the education? What are possible reasons accounting for the patterns you see?
```{r}
levels(dat$marital)
dat$marital<-factor(dat$marital, levels=(c("Married","Never married","Divorced","Widowed","Separated","No answer")))
ggplot(dat, aes(x=marital, y=education))+geom_boxplot()+ggtitle("Marital Status vs. Education Level")+xlab('Marital Status')+ylab('Education Level')
AvgEduInMar<-dat%>%
  group_by(marital)%>%
  summarize(mean_education=mean(education,na.rm=TRUE))%>%
  arrange(mean_education)
AvgEduInMar
ggplot(AvgEduInMar, aes(x=marital, weight=mean_education))+geom_bar()+coord_flip()+ggtitle("Mean Education Year by Marital Status")+ylab("Highest Year of Education (mean)") +xlab("Marital Status")
dat%>%
  select(marital, education)%>%
  group_by(marital)%>%
  count(education)
```
**As shown by the graphs above, there is no clear correlation between marital status and education level. They all are around the same.**